Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-08-22 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added background text

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

>>>>

\Subsection{The Fire Investigators' Challenge}

For the moment lets step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occured in area of older townhomes.  The socio-economics of the neighboorhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene, "all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The fornt door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and chaild were rescued from a bedroom upstairs".  You thank the fire officier and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his work place.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit, townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  The is some thermal damage and smoke deposits on the from of the siding on the porch.  There outer pane of the living room window is for the most part still intact. you notice that three of the four windows on the second floor have been removed.  On the "D side" of the townhome there are no openings and no smoke or fire damage is present.  On the rear or "Side C"
of the house the door and windows are intact but open.  There is no somke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home

You take some notes and photograph the exterior of the structure, the fire department is not repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage.  Looking to your right, two chairs infront of the living room window have some thermal damage.  There are a clear horizontal line of demarcation the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the "D wall" of the living room was a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is not thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an uphostered sofa and chair.  The wall behind the sofa and chair has extensive thermal danage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  Their is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  THe fire damage doe not extend beyond the top of the stairs.  there is smoke damage in the upstairs hallway an in the front bedrooms. THE FURTHER YOU MOVE FROM THE stairs the less damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focussing on the area with the most damage, the area with the sofa and chair.  As you are looking toward 'b wall' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  The corner formed by the wall behind and the wall of the stairway has a "V pattern" burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The wire from the heater extends toward the wall outlet behind the sofa.
You have no developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on and left until it ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine of the heater switch was in the "on" position.  No you need to start testing your hypothesis.  Is everything consistant with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the rightside back cushion of the sofa.
the damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee looks like a arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the soaf frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair clearly has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not be moved.  The cushion was heated from the top down.  Perhaps the constrion of the chair slowed the flame spread through the chair from the bottom up, if the heatere was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews whatelse can you do to eliminate the inconsistancies?   We will reexamine the in Chapter XX.

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Associaand tion, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.
